---
title: 'Millennials: Changing the Face of Political Participation in Germany?'
author: "Philippe Joly and Marcus Spittler"
date: "6/28/2019"
output:
  pdf_document:
    dev: cairo_pdf
  word_document:
    reference_docx: word-style.docx
  html_document:
    df_print: paged
nocite: |
  @dalton_good_2015, @grasso_generations_2016, @norris_democratic_2002
bibliography: ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..")
```

```{r load_packages, include=FALSE}
library(dplyr) # Data wrangling
library(ggplot2) # Graphs
library(kableExtra) # Export tables
```

```{r load_functions, include=FALSE}
pct <- function(df, generation, action, digit_level = 0){
  df[df$generation == generation, action] %>%
    round(digit_level) %>%
    as.character() %>%
    paste0("%")
}
```

```{r load_data, include=FALSE}
ess <- readRDS("../data/ess_allctries_19.rds")
```

```{r sample_characteristics, include=FALSE}
samplesize <- ess %>%
  filter(!is.na(generation)) %>%
  count()  %>%
  as.integer()

yrbrn_min <- ess %>%
  filter(!is.na(generation)) %>%
  .$yrbrn %>%
  min()

yrbrn_max <- ess %>%
  filter(!is.na(generation)) %>%
  .$yrbrn %>%
  max()
```

> Millenials, born between 1985 and 2000, are often critized for being self-serving and little interested in politics. We examine the political participation of the Millenial generation in comparison to Generation X and Boomers in Germany. We find that Millennials participate today more than the Generation X at the age of twenty and, if the current trend persists, they will participate more than Boomers did during their thirties. Therefore, Millenials might actually contribute to an increasing level of participation in Germany.

# Introduction

Millenials are often critized for being self-serving and little interested in politics. The Time Magazine even called them the "Me Me Me Generation". Over the recent years, however, Millennials, born between 1985 and 2000, have proven the opposite. Not only digital politics, but also climate protection brings numerous young people every week to the streets under the motto "Fridays for Future", and this despite considerable criticism from older people. The last European elections have also shown a considerable gap between the political preferences of younger and older age groups, especially in Germany. These developments have sparked a new interest – in academia and in the wider public alike – for the political participation of young citizens.

In this article we compare the political participation of Millennials with those of Baby boomers (born between 1955 and 1969) and the Generation X (_Xers_, born between 1970 and 1984) in Germany while taking into account the methodological difficulties associated with this type of research. We examine how exactly Millennials are engaging in politics and whether they are participating more or less in comparison to older generations.

# Changing Political Participation in Western Europe

To understand how Millennials get involve into politics, we need to take a step back and consider long-term dynamics of political participation. According to the literature, two important trends have characterized the evolution of political participation in Western Europe over the last decades. First, we observe _a decline in voter turnout_ since the Second World War. The decline is affecting all age groups, but even more so the youngest ones. In Germany, voter turnout for federal elections reached its peak in 1972, at 91.1% (in West Germany), and its lowest point in 2009, at 70.8%. Second, data from repeated, nationally-representative surveys suggest there has been _a rise in nonelectoral forms of participation_ since the 1970s. These forms of participation include activities such as signing petitions, joining in boycotts, or taking part in demonstrations. These political actions are not directly linked to the electoral process (although they might be). They are more spontaneous and less dependent on attachment to hierarchical organizations like parties or labor unions.

Scholars have been debating whether these trends are attributable to period or cohort effects. Period effects might result from technological evolution, a growing acceptance by authorities for alternative forms of participation, a greater institutionalization of social movement organizations, and transformations of post-industrial economies characterized by an expanding service sector and the diffusion of post-secondary education. Cohort (or generational) effects might result from processes of political socialization through which young citizens would have adopted a distinct set of attitudes and developed particular ways of participating in politics. Some authors argue that younger, postmaterialistic cohorts increasingly value autonomy and self-expression, which leads them to embrace elite-challenging forms of participation.

Unfortunately, we might never be able to determine with certainty what from period or cohort effects is driving these long-term trends in political participation. Indeed, separating age, period, and cohort (APC) effects poses a statistical problem: age is perfectly explained by cohort and period (simply subtract the year of birth from the measurement year). No statistical solution allows for the perfect separation o f the age, period, and cohort effects. Yet, this does not mean that scholars should completely avoid comparing the political behavior of generations. By constraining one of the three terms of the APC equation, informed estimations about the nature of social change can still be made.

Despite the inherent challenges associated with generational analysis, it is surprising that so few studies have examined the political participation of Millennials. Most studies that have looked at the decline in voter turnout and the expansion of nonelectoral participation were conducted in the 1990s and beginning of the 2000s. Citizens of Generation X, not the Millennials, were the “young” at that time. How are Millennials participating in politics? Are they simply pursuing the same old trends observed before; voting less, but getting involved in nonelectoral forms of political participation?

# Data

We look here at the volume and modes of participation of Millennials compared to Baby boomers and the Generation X in Germany. We begin by exploring trends in voter turnout by age groups using data from the _Bundeswahlleiter_. We then examine the three generations' involvement in other forms of electoral and non-electoral political participation using data from eight biennial rounds of the European Social Survey (ESS) conducted between 2002 and 2017 [1].The ESS asked respondents whether they had engaged in various forms of political participation in the 12 months preceding each of the survey rounds. We compare how Boomers, Xers, and Millenials participate in two electoral activities -- contacting a politician or a government official and working in a political -- and four non-electoral activities -- working in another organisation or association, signing a petition, taking part in a lawful demonstration, and boycotting certain products.

# A Distinct Trajectory of Political Participation?

```{r collapse, include=FALSE}
ess_gr <- ess %>%
  filter(!is.na(generation)) %>%
  filter(age_doi >= 16) %>%
  group_by(generation) %>%
  summarize_at(vars(polintr,
                    vote,
                    contplt,
                    wrkprty,
                    wrkorg,
                    badge,
                    sgnptit,
                    pbldmn,
                    bctprd,
                    clsprty),
               ~weighted.mean(., dweight,  na.rm = T )*100)
```


```{r export_table_averages, echo=FALSE}
# ess_gr %>%
#   kable() %>%
#   kable_styling()
```

In Germany, Millennials became eligible to vote in a federal election for the first time in 2005. By then, the oldest members of this generation were `r 2005 - 1985` years old. In the subsequent elections, they tended to show up less to the polls than older age groups according to data from the Bundeswahlleiter. This pattern, however, seems to reflect more a life-course than a generational effect. Scholars usually explain this tendency by the fact that young people have a looser connection to their community. They are often still in education and are more mobile. The process of "transitioning to adulthood," that is, settling in a community, working, cohabiting with a partner, and raising a family, is associated with an increase in electoral participation. An interesting development in Germany has been the recent rise in voter turnout. After reaching its lowest in point in 2009, voter turnout has increased to 71.5% in 2013 and to 76.2% in 2017. Remarkably, the _gap_ between the most involved age group (60 to 69 years old) and the least involved one (21 to 24 years old) declined from `r 79.2 - 58.6`%  in 2009 to `r 81.0 - 67.0`% in 2017. The increase in voter turnout among the youngest age groups (18 to 34 years old) was 2% to 3% higher than the overall 5.4% increase since 2009. Millennials, like a rising tide, seem to have partially catched-up with older cohorts.

Turning to the other forms of political participation measured by the ESS from 2002 to 2017, Millennials' involvement varies depending on the type of activity. They were less likely to contact a politician (`r pct(ess_gr, "Millennial", "contplt")` of them did so compared to `r pct(ess_gr, "Xer", "contplt")` for Xers and `r pct(ess_gr, "Boomer", "contplt")` for Boomers), less likely to sign a petition (`r pct(ess_gr, "Millennial", "sgnptit")` compared to `r pct(ess_gr, "Xer", "sgnptit")` for Xers and `r pct(ess_gr, "Boomer", "sgnptit")` for Boomers), and less likely to boycott certain products (`r pct(ess_gr, "Millennial", "bctprd")` compared to `r pct(ess_gr, "Xer", "bctprd")` for Xers and `r pct(ess_gr, "Boomer", "bctprd")` for Boomers). Millennials, however, were more inclined to work for a party (`r pct(ess_gr, "Millennial", "wrkprty", digit_level = 1)`) and work for an organization (`r pct(ess_gr, "Millennial", "wrkorg")`) than Xers (respectively, `r pct(ess_gr, "Xer", "wrkprty", digit_level = 1)` and `r pct(ess_gr, "Xer", "wrkorg")`), but still less inclined to do so than Boomers (respectively, `r pct(ess_gr, "Boomer", "wrkprty")` and `r pct(ess_gr, "Boomer", "wrkorg")`). Millennials' participation in demonstrations (`r pct(ess_gr, "Millennial", "pbldmn")`) exceeded those of the other two generations (`r pct(ess_gr, "Xer", "pbldmn")` for Xers and `r pct(ess_gr, "Boomer", "pbldmn")` for Boomers). Looking at the results, we cannot conclude that Millennials are particularly focused on nonelectoral forms of political participation. Except for demonstrations, their participation is rather moderate and balanced across different political actions.

As for voter turnout, the relatively low levels of participation of Millennials might reflect a life-course rather than a generational effect. Since the ESS covers 16 years, we can use this range to compare the participation of respondents of the three generations as a function of their age at the date of the survey interview (see _Figure_). Here, the level of participation is simply measured as the predicted number of political activities performed by the respondents in the year preceding the survey. Respondents can score up to a maximum of six points when they took part in the two electoral and four nonelectoral forms of political participation presented previously.

All three generations tend to participate more as they get older, but each of them appears to follow a distinct trajectory. Interestingly, we find that Millennials' overall participation is high for their age and even _surpasses_ Xers' participation in their twenties. It is still too early to argue that this trend is caused by a generational effect. Millennials high involvement might simply echo the highly politicized period in which they gew in. The economic crisis, the flow of refugees, climate change, and the rise of the radical right seem to have reactivated the political participation of all cohorts in Germany. We will have to continue tracking Millennials' participation to see whether they maintain their distinct trajectory in the future. As of now, we can nonetheless safely conclude that, contrary to some common belief, Millennials are not an apathetic generation.


```{r generate_graph, echo=FALSE, message=F}
# Load WZB Dahrendorf font into R - Install font in Windows first
if(Sys.info() %>% {stringr::str_detect(tolower(.), "windows" )} %>% any()){
  windowsFonts(dahrendorf = windowsFont("Dahrendorf"))
}

# Add translation and set language
giveText <- function( whichText, language = "EN"){
  switch(language,
    EN = switch(whichText,
      title = "Political participation by generation and age",
      x = "Age at date of interview",
      y = "Predicted number of political activities",
      source = "Source: European Social Survey (www.europeansocialsurvey.org)"
    ),
    DE = switch(whichText,
      title = "Politische Beteiligung nach Generation und Alter",
      x = "Alter zum Befragungszeitpunkt",
      y = "Vorhergesagte Anzahl politischer Aktivitäten",
      source = "Quelle: European Social Survey (www.europeansocialsurvey.org)"
    )
  )
}
language <- "EN" # or "DE"

# Sum number of activities
ess %>%
  filter(complete.cases(generation,
                        contplt,
                        wrkprty,
                        wrkorg,
                        sgnptit,
                        pbldmn,
                        bctprd)) %>%
  mutate(sumpolpart = contplt + wrkprty + wrkorg + sgnptit + pbldmn +
         bctprd) %>%
  mutate(generation = factor(generation,
                             levels=rev(levels(generation)))) %>%
  filter(age_doi >= 16) %>%                                                      
  ggplot(aes(
    x = age_doi, y = sumpolpart, group = generation, color = generation, weight = dweight)) +
  geom_smooth(method = 'loess', formula = 'y ~ x') +
  ggtitle(giveText("title", language)) +
  labs(x = giveText("x", language),
       y = giveText("y", language),
       caption = giveText("source", language)) +
  scale_x_continuous(breaks = c(16, seq(20,65,5))) +
  scale_y_continuous(
    breaks = seq(0,2,.5),
    labels = scales::comma(seq(0,2,.5), accuracy = .1, big.mark = "",
      decimal.mark = ifelse(language == "DE", ",", "."))) +
  scale_color_manual(
    values = c("#c7007e", "#f39100", "#0081b6"),
    name = "",
    labels = c("Millennials (1985-2000)", # Should this not be 2000?
               "Generation X (1970-84)",
               "Boomers (1955-69)")
  ) +
  coord_cartesian(ylim = c(0,2), expand = TRUE) +
  theme(
    text = element_text(family = "dahrendorf"),
    plot.background = element_rect(color = "#d9dada", fill = "#d9dada"),
    panel.background = element_rect(fill = "#d9dada"),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    legend.background = element_rect(fill = "#d9dada"),
    legend.key = element_rect(color = "#d9dada"),
    legend.position = "bottom"
  )

ggsave(paste0("text/number_activities_age_doi_", language,".pdf"),
       device = cairo_pdf, width = 19, height = 19/1.35,
       dpi = 600, units = "cm")

```

# Conclusions

Our analysis has shown that Millennials are, in general, less active than the older generations, but this lower level of engagement might be an artifact of an age effect: On average, we can say that Millennials participate today more than Xers during their twenties and, if the current trend persists, they will participate more than Boomers did during their thirties. Millennials are in fact more mobilized than Xers were at the same age and do not participate in a radically different way. Therefore, they might actually contribute to an increase in the overall level of participation in Germany. That’s good news for democracy, as political participation is a clear indicator that the input-dimension of the political system is functioning. High levels of participation also improve democratic quality and the legitimacy a democracy can claim for itself.

Two aspects were left open in this contribution: First, how did the invention of the World Wide Web affect the political repertoire of Millennials in comparison to older generations, and is online participation supplementing or substituting other forms of participation? Second, what will happen with Millennials' political engagement when they transition further into adulthood, settle down in stable jobs, pay taxes, raise families, and interact more with the welfare state. Will they keep on participating at this level?

# Notes
1. The ESS 1 fieldwork was conducted between 20.11.02 and 16.05.03; the ESS 8 fieldwork, between 23.08.16 and 26.03.17. We use a sample of `r samplesize` respondents born between `r yrbrn_min` and `r yrbrn_max`.
2. See Bundeswahlleiter, 2018, "Wahl zum 19. deutschen Bundestag am 24. September 2017. Heft 4: Wahlbeteiligung und Stimmabgabe der Frauen und Männer nach Altersgruppen", https://www.bundeswahlleiter.de/dam/jcr/e0d2b01f-32ff-40f0-ba9f-50b5f761bb22/btw17_heft4.pdf

# Authors

_Philippe Joly_ is a research fellow in the research unit Democracy and Democratization and a doctoral candidate at Humboldt University. His areas of specialization are political participation, social movements, and democratization. His PhD thesis examines the effect of political socialization on the development of political protest in new democracies, especially in the context of post-communist transformations.

_Marcus Spittler_ is a research fellow in the research unit Democracy and Democratization since 2015. Before he worked at the Electoral Integrity Project at the University of Sydney. His areas of research are in comparative politics and electoral behavior, while his PhD thesis focuses on the effect of populism on Democracy from an empirical perspective.

# Further reading
