---
title: "Youth Political Participation"
author: "Philippe Joly and Marcus Spittler"
date: "6/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..")
```

```{r load_packages, message = FALSE}
library(dplyr) # Data wrangling
library(ggplot2) # Graphs
library(kableExtra) # Export tables
```

```{r load_data}
ess <- readRDS("data/ess.rds")
```

```{r load_functions}
percent <- function(df, generation, action){
  df[df$generation == generation, action] %>% 
    round() %>% 
    as.character() %>% 
    paste0("%")
}
```

# Millennials: Changing the Face of Political Participation in Germany?

## Outline

- Introduction [1000]
    - Friday4Future shows the youth engagement in non-representational activities. General wisdom: We see a rise of that.
- Literature review [2500]
    - Puzzle: So if we see such a rise of non-representational participation, is 1) age, cohort, period effect? and 2) an extension of the existing repertoire or its replacement?
    - Methodological problems explained and how we tackle them here in an "age at date of interview"-plot
- Empirical findings [5000]
    - Representational
        - Feel closer to a party [plot here?]
        - Voting
    - Non-representational
        - Political interest [2000]
        - Worked in another organization
        - Boycotts
- Summary [1000]


## Introduction

- The last European elections and the Friday for Future demonstrations have sparked a new interest – in academia and in the wider public – for the political participation of young citizens.
- Questions:
    - But how exactly are Millennials engaging in politics? Are they participating more or less than older generations? Does their repertoire of participation differ significantly?
- In this paper:
    - We review the literature on generations and political participation. We highlight the very tricky data and methodological problems associated with this type of research.
    - We compare the repertoire of political participation of Millennials with those of Baby boomers and Xers in Germany using data of the European Social Survey collected from 2002 [2003?] to 2016 [2017?].
    - We show that Millennials are in general less active than the older generations, but this lower level of engagement might be an artifact of an age effect: Millennials are sometimes more mobilized, in certain forms of political participation, than Xers were at the same age.

## Millenials and changing political participation in Western Europe

To understand how Millennials get involve into politics, we need to take a step back and consider long-term dynamics of political participation. According to the literature, two important trends have characterized the evolution of political participation in Western Europe over the last decades. First, we observe **a decline in voter turnout** since the Second World War. The decline is affecting all age groups, but even more so the youngest ones. In Germany, voter turnout for federal elections reached its peak in 1972, at 91.1% (in West Germany), and its lowest point in 2009, at 70.8%. Second, data from repeated, nationally-representative surveys suggest there has been **a rise in non-electoral forms of participation** since the 1970s. These forms of participation include activities such as signing petitions, joining in boycotts, or taking part in demonstrations. These political actions are not directly linked to the electoral process (although they might be). They are more spontaneous and less dependent on attachment to hierarchical organizations like parties or labor unions.

Scholars have been debating whether these trends are attributable to period or cohort effects. Period effects might result from technological evolution, a growing acceptance by authorities for alternative forms of participation, a greater institutionalization of social movement organizations, and transformations of post-industrial economies characterized by an expanding service sector and the diffusion of post-secondary education. Cohort (or generational) effects might result from processes of political socialization through which young citizens would have adopted a distinct set of attitudes and developed particular ways of participating in politics. Some authors argue that younger, postmaterialistic cohorts increasingly value autonomy and self-expression, which leads them to embrace elite-challenging forms of participation. 

Unfortunately, we might never be able to determine with certainty what from period or cohort effects is driving these long-term trends in political participation. Indeed, separating age, period, and cohort (APC) effects poses a statistical problem: age is perfectly explained by cohort and period (simply subtract the year of birth from the measurement year). No statistical solution allows for the perfect separation of the age, period, and cohort effects. Yet, this does not mean that scholars should completely avoid comparing the political behavior of generations. By constraining one of the three terms of the APC equation, informed estimations about the nature of social change can still be made.

Despite the inherent challenges associated with generational analysis, it is surprising that so few studies have examined the political participation of Millennials. Most studies that have looked at the decline in voter turnout and the expansion of non-electoral participation were conducted in the 1990s and beginning of the 2000s. Citizens of Generation X, not the Millennials, were the “young” at that time. How are Millennials participating in politics? Are they simply pursuing the same old trends observed before; voting less, but getting involved in non-electoral forms of political participation? Or have they developped their own distinct model of political activism? Finally, what is the participation of Millennials telling us about the possible future of representative democracy?


## Methods

In this article, we look at the volume and modes of participation of Millennials (born after 1985) compared to Baby boomers (born between 1955 and 1969) and the Generation X (born between 1970 and 1984) in Germany. 

We begin by exploring trends in voter turnout by age groups using data from the _Bundeswahlleiter_. 

We then examine the three generations' involvement in other forms of electoral and non-electoral political participation using data from eight biennial rounds of the European Social Survey (ESS) conducted between 2002 and 2017 [1].The ESS asked respondents whether they had engaged in various forms of political participation in the 12 months preceding each of the survey rounds. We compare how Boomers, Xers, and Millenials participate in two electoral activities -- 1) contacting a politician or a government official and 2) working in a political -- and four nonelectoral activities -- 1) working in another organisation or association, 2) signing a petition, 3) taking part in a lawful demonstration, and 4) boycotting certain products. 

```{r sample_characteristics}
samplesize <- ess %>% 
  filter(!is.na(generation)) %>%
  count()  %>%
  as.integer()

yrbrn_min <- ess %>% 
  filter(!is.na(generation)) %>%
  .$yrbrn %>%
  min()

yrbrn_max <- ess %>% 
  filter(!is.na(generation)) %>%
  .$yrbrn %>%
  max()
```


## Findings

```{r collapse}
ess_gr <- ess %>%
  filter(!is.na(generation)) %>%
  group_by(generation) %>%
  summarize_at(vars(polintr,
                    vote,
                    contplt,
                    wrkprty,
                    wrkorg,
                    badge,
                    sgnptit,
                    pbldmn,
                    bctprd,
                    clsprty), 
               ~weighted.mean(., dweight,  na.rm = T )*100)
```


```{r export_table_averages}
ess_gr %>% 
  kable() %>%
  kable_styling()
```

In Germany, Millennials became eligible to vote in a federal election for the first time in 2005. By then, the oldest members of this generation were `r 2005 - 1985` years old. In the subsequent elections, they tended to show up less to polls than older age groups according to data from the Bundeswahlleiter. This pattern, however, seems to reflect more a life-course than a generational effect. Scholars usually explain this tendency by the fact that young people have a loser connection to their community: They are often still in education and are more mobile. The process of "transitioning to adulthood," that is, settling in a community, working, cohabiting with a partner, and raising a familly, is associated with an increase in electoral participation. An interesting development in Germany has been the recent rise in voter turnout. After reaching its lowest in point in 2009, voter turnout has increased to 71.5% in 2013 and to 76.2% in 2017. Remarkably, the _gap_ between the most involved age group (60 to 69 years old) and the least involved one (21 to 24 years old) declined from `r 79.2 - 58.6`%  in 2009 to `r 81.0 - 67.0`% in 2017. The increase in voter turnout among the youngest age groups (18 to 34 years old) was 2% to 3% higher than the overall 5.4% increase since 2009. Millennials, like a rising tide, seem to have catched-up some of the gap that separated them from older cohorts. 

Turning to the other forms of political participation measured by the ESS from 2002 to 2017, Millennials' involvement varies depending on the type of activity. They were less likely to contact a politician (`r percent(ess_gr, "Millennial", "contplt")` of them did so compared to `r percent(ess_gr, "Xer", "contplt")` for Xers and `r percent(ess_gr, "Boomer", "contplt")` for Boomers), less likely to sign a petition (`r percent(ess_gr, "Millennial", "sgnptit")` compared to `r percent(ess_gr, "Xer", "sgnptit")` for Xers and `r percent(ess_gr, "Boomer", "sgnptit")` for Boomers), and less likely to boycott certain products (`r percent(ess_gr, "Millennial", "bctprd")` compared to `r percent(ess_gr, "Xer", "bctprd")` for Xers and `r percent(ess_gr, "Boomer", "bctprd")` for Boomers). Millennials, however, were more inclined to work for a party (`r percent(ess_gr, "Millennial", "wrkprty")`) and work for an organization (`r percent(ess_gr, "Millennial", "wrkorg")`) than Xers (respectively, `r percent(ess_gr, "Xer", "wrkprty")` and `r percent(ess_gr, "Xer", "wrkorg")`), but still less inclined to do so than Boomers (respectively, `r percent(ess_gr, "Boomer", "wrkprty")` and `r percent(ess_gr, "Boomer", "wrkorg")`). Millennials' participation in demonstrations (`r percent(ess_gr, "Millennial", "pbldmn")`) exceeded those of the other two generations (`r percent(ess_gr, "Xer", "pbldmn")` for Xers and `r percent(ess_gr, "Boomer", "pbldmn")` for Boomers). Looking at the results, we cannot conclude that Millennials are particularly focused on non-electoral forms of political participation. Except for demonstrations, their participation is rather moderate and balanced accross different political actions. 

As for voter turnout, the relatively low levels of participation of Millennials might reflect a life-course rather than a generational effect. Since the ESS covers 16 years, we can use this range to compare the participation of respondents of the three generations as a function of their age at the date of the survey interview (see _Figure_). Here, the level of participation is simply measured as the predicted number of political activities performed by the respondents in the year preceding the survey. Respondents can score up to a maximum of 6 when they took part in the two electoral and in the four non-electoral forms of political participation presented previously. 

All three generations tend to participate more as they get older, but each of them appears to follow a distinct trajectory. Interestingly, we find that Millennials' overall participation is high for their age and even surpasses Xers' participation in their twenties. It is still too early to argue that this trend is caused by a generational effect. Millennials high involvement might simply echo the highly politicized period in which they gew in. The economic crisis, the flow of refugees, climate change, and the rise of the radical right seem to have reactivated the political participation of all cohorts in Germany. We will have to continue tracking Millennials' participation to see if they maintain their distinct trajectory in the future. As of now, we can nonetheless safely conclude that fears that Millennials are apathetic are unwarranted. 


```{r generate_graph}

# Sum number of activities
ess %>%
  filter(complete.cases(generation, 
                        contplt, 
                        wrkprty, 
                        wrkorg, 
                        sgnptit, 
                        pbldmn,
                        bctprd)) %>%
  mutate(sumpolpart = contplt + wrkprty + wrkorg + sgnptit + pbldmn +
         bctprd) %>%
  ggplot(aes(
    x = age_doi, y = sumpolpart, group = generation, color = generation, weight = dweight)) + 
  geom_smooth(method = 'loess', formula = 'y ~ x') +
  labs(x = "Age at date of interview",
       y = "Predicted number of political activities") +
  scale_color_discrete(name="",
                     labels = c("Baby boomers (1955-69)", 
                                "Generation X (1970-84)", 
                                "Millennials (1985-...)"))

```

[According to the graph, Millennials appear to be more active than Xer at the same age].

## Conclusion

- [Normative considerations] Democracies benefit from engaged citizens. Political participation helps to monitor the activities of governing elites and improves political representation.
- Recently, many people have been wondering whether Millennials would change the face of political participation in Germany.
- Our results show that Millennials do not participate in a radically different way than Baby boomers and Xers.
- They are usually a bit less active, except in demonstrations.
- Yet, these results might be linked to an age effect.
- Millennials are actually more active in politics than Xers at the same age.
- Therefore, if they continue on this trajectory, they might actually contribute to an increase in the overall level of participation in Germany.
- That’s good news at the time when the republic is faced with many challenges [list some].
- It will be interesting to follow the political participation of Millennials as they transition to adulthood: settle down in stable jobs, pay taxes, raise families, and interact more with the welfare state.

## Further reading



## Notes
1. The ESS 1 fieldwork was conducted between 20.11.02 and 16.05.03; the ESS 8 fieldwork, between 23.08.16 and 26.03.17. We use a sample of `r samplesize` respondents born between `r yrbrn_min` and `r yrbrn_max`. 
2. See https://www.bundeswahlleiter.de/dam/jcr/e0d2b01f-32ff-40f0-ba9f-50b5f761bb22/btw17_heft4.pdf 
